# Data wrangling with R

This code is based on the original work from [HealthyR](https://healthyr.surgicalinformatics.org) and has been adapted by **Laura Bravo**

### Tidyverse packages: ggplot2, dplyr, tidyr, etc.

Most of the functions introduced in this session come from the [tidyverse family](http://tidyverse.org/), rather than Base R. Including `library(tidyverse)` in your script loads a list of packages: ggplot2, dplyr, tidyr, forcats, etc.

```{r}
library(tidyverse)
setwd("/Users/l.bravo@bham.ac.uk/Library/CloudStorage/OneDrive-UniversityofBirmingham/Desktop/ML_Classes2025/Day1/Practical")
library(here)
```



In this session we will get to know key functions to work with data:

-   summarise data using: `group_by()`, `summarise()`, and `mutate()`;
-   reshape data between the wide and long formats: `pivot_wider()` and `pivot_longer()`;
-   `select()` columns and `arrange()` (sort) rows.

## Data

We are using a condensed version of the [Global Burden of Disease Data](https://ghdx.healthdata.org) dataset.

```{r}
#Can load the dataset through many ways but quickest and one i always resort to, is data.table function
library(data.table)
gbd_full <- fread("global_burden_disease_cause-year-sex-income.csv")
```

```{r}
head(gbd_full)
```

```{r}

# Creating a single-year tibble for printing and simple examples:
gbd2017 <- gbd_full %>% 
  filter(year == 2017)

```

## Plot the data

The best way to investigate a dataset is to plot it (do not worry if do not understand the code, will do a separate exercise on plotting later)

```{r}
gbd2017 %>% 
  # without the mutate(... = fct_relevel()) 
  # the panels get ordered alphabetically
  mutate(income = fct_relevel(income,
                              "Low",
                              "Lower-Middle",
                              "Upper-Middle",
                              "High")) %>% 
  # defining the variables using ggplot(aes(...)):
  ggplot(aes(x = sex, y = deaths_millions, fill = cause)) +
  # type of geom to be used: column (that's a type of barplot):
  geom_col(position = "dodge") +
  # facets for the income groups:
  facet_wrap(~income, ncol = 4) +
  # move the legend to the top of the plot (default is "right"):
  theme(legend.position = "top")
```

## Aggregating: `group_by()`, `summarise()`

Health data analysis is frequently concerned with making comparisons between groups. Groups of genes, or diseases, or patients, or populations, etc. An easy approach to the comparison of data by a categorical grouping is therefore essential.

We will introduce flexible functions from **tidyverse** that you can apply in any setting.

To quickly calculate the total number of deaths in 2017, we can select the column and send it into the `sum()` function:

```{r}
gbd2017$deaths_millions %>% sum()
```

But a much cleverer way of summarising data is using the `summarise()` function:

```{r}
gbd2017 %>% 
  summarise(sum(deaths_millions))
```

`sum()` is a function that adds numbers together, whereas `summarise()` is an efficient way of creating summarised dataframes. The main strength of `summarise()` is how it works with the `group_by()` function. `group_by()` and `summarise()` are a perfect complement for each other and used togetehr most of the time.

We use `group_by()` to tell `summarise()` which subgroups to apply the calculations on. In the above example, without `group_by()`, summarise just works on the whole dataset, yielding the same result as just sending a single column into the `sum()` function.

We can subset on the cause variable using `group_by()`:

```{r}
gbd2017 %>% 
  group_by(cause) %>% 
  summarise(sum(deaths_millions))
```

Furthermore, `group_by()` is happy to accept multiple grouping variables. So by just copying and editing the above code, we can quickly get summarised totals across multiple grouping variables (by just adding `sex` inside the `group_by()` after `cause`):

```{r}
gbd2017 %>% 
  group_by(cause, sex) %>% 
  summarise(sum(deaths_millions))
```

## Add new columns: `mutate()`

Let's first give the summarised column a better name, e.g., `deaths_per_group`. We can remove groupings by using `ungroup()`. This is important to remember if you want to manipulate the dataset in its original format. We can combine `ungroup()` with `mutate()` to add a total deaths column, which will be used below to calculate a percentage:

```{r}
gbd2017 %>% 
  group_by(cause, sex) %>% 
  summarise(deaths_per_group = sum(deaths_millions)) %>% 
  ungroup() %>% 
  mutate(deaths_total = sum(deaths_per_group))
```

So `summarise()` condenses a dataframe, whereas `mutate()` retains its current size and adds columns.

### Percentages formatting: `percent()`

We can also add further lines to `mutate()` to calculate the percentage of each group:

```{r, message = FALSE}
# percent() function for formatting percentages come from library(scales)
library(scales)
gbd2017_summarised <- gbd2017 %>% 
  group_by(cause, sex) %>% 
  summarise(deaths_per_group = sum(deaths_millions)) %>% 
  ungroup() %>% 
  mutate(deaths_total    = sum(deaths_per_group),
        deaths_relative = deaths_per_group/deaths_total)
gbd2017_summarised
```

## `summarise()` vs `mutate()`

So far we've shown you examples of using `summarise()` on grouped data (following `group_by()`) and `mutate()` on the whole dataset (without using `group_by()`).

But here's the thing: `mutate()` is also happy to work on grouped data.

Let's save the aggregated example from above in a new dataframe. We will then sort the rows using `arrange()` based on `sex`, just for easier viewing (it was previously sorted by `cause`).

The `arrange()` function sorts the rows within a tibble:

```{r}
gbd_summarised <- gbd2017 %>% 
  group_by(cause, sex) %>% 
  summarise(deaths_per_group = sum(deaths_millions)) %>% 
  arrange(sex)

gbd_summarised
```

You should also notice that `summarise()` drops all variables that are not listed in `group_by()` or created inside it. So `year`, `income`, and `deaths_millions` exist in `gbd2017`, but they do not exist in `gbd_summarised`.

We now want to calculate the percentage of deaths from each cause for each gender. We could use `summarise()` to calculate the totals:

```{r}
gbd_summarised_sex <- gbd_summarised %>% 
  group_by(sex) %>% 
  summarise(deaths_per_sex = sum(deaths_per_group))

gbd_summarised_sex
```

But that drops the `cause` and `deaths_per_group` columns. One way would be to now use a join on `gbd_summarised` and `gbd_summarised_sex`:

```{r}
full_join(gbd_summarised, gbd_summarised_sex)
```

Joining different summaries together can be useful, especially if the individual pipelines are quite long (e.g., over 5 lines of `%>%`). However, it does increase the chance of mistakes creeping in and is best avoided if possible.

An alternative is to use `mutate()` with `group_by()` to achieve the same result as the `full_join()` above:

```{r}
gbd_summarised %>% 
  group_by(sex) %>% 
  mutate(deaths_per_sex = sum(deaths_per_group))
```

So `mutate()` calculates the sums within each grouping variable (in this example just `group_by(sex)`) and puts the results in a new column without condensing the tibble down or removing any of the existing columns.

Let's combine all of this together into a single pipeline and calculate the percentages per cause for each gender:

```{r}
gbd2017 %>% 
  group_by(cause, sex) %>% 
  summarise(deaths_per_group = sum(deaths_millions)) %>% 
  group_by(sex) %>% 
  mutate(deaths_per_sex  = sum(deaths_per_group),
         sex_cause_perc = percent(deaths_per_group/deaths_per_sex)) %>% 
  arrange(sex, deaths_per_group)
```

## Common arithmetic functions - `sum()`, `mean()`, `median()`, etc.

Statistics is an R strength, so if there is an arithmetic function you can think of, it probably exists in R.

The most common ones are:

-   `sum()`
-   `mean()`
-   `median()`
-   `min()`, `max()`
-   `sd()` - standard deviation
-   `IQR()` - interquartile range

An import thing to remember relates to missing data: if any of your values is NA (not available; missing), these functions will return an NA. Either deal with your missing values beforehand (recommended) or add the `na.rm = TRUE` argument into any of the functions to ask R to ignore missing values.

```{r}
mynumbers <- c(1, 2, NA)
sum(mynumbers)
sum(mynumbers, na.rm = TRUE)
```

## `select()` columns

The `select()` function can be used to choose, rename, or reorder columns of a tibble.

For the following `select()` examples, let's create a new tibble called `gbd_2rows` by taking the first 2 rows of `gbd_full` (just for shorter printing):

```{r}
gbd_2rows <- gbd_full %>% 
  slice(1:2)

gbd_2rows
```

Let's `select()` two of these columns:

```{r}
gbd_2rows %>% 
  select(cause, deaths_millions)
```

We can also use `select()` to rename the columns we are choosing:

```{r}
gbd_2rows %>% 
  select(cause, deaths = deaths_millions)
```

The function `rename()` is similar to `select()`, but it keeps all variables whereas `select()` only kept the ones we mentioned:

```{r}
gbd_2rows %>% 
  rename(deaths = deaths_millions)
```

`select()` can also be used to reorder the columns in your tibble. Moving columns around is not relevant in data analysis (as any of the functions we showed you above, as well as plotting, only look at the column names, and not their positions in the tibble), but it is useful for organising your tibble for easier viewing.

So if we use select like this:

```{r}
gbd_2rows %>% 
  select(year, sex, income, cause, deaths_millions)
```

The columns are reordered.

If you want to move specific column(s) to the front of the tibble, do:

```{r}
gbd_2rows %>% 
  select(year, sex, everything())
```

And this is where the true power of `select()` starts to come out. In addition to listing the columns explicitly (e.g., `mydata %>% select(year, cause...)`) there are several special functions that can be used inside `select()`. These special functions are called select helpers, and the first select helper we used is `everything()`.

The most common select helpers are `starts_with()`, `ends_with()`, `contains()`, `matches()` (but there are several others that may be useful to you, so press F1 on `select()` for a full list, or search the web for more examples).

Let's say you can't remember whether the deaths column was called `deaths_millions` or just `deaths` or `deaths_mil`, or maybe there are other columns that include the word "deaths" that you want to `select()`:

```{r}
gbd_2rows %>% 
  select(starts_with("deaths"))
```

Note how "deaths" needs to be quoted inside `starts_with()` - as it's a word to look for, not the real name of a column/variable.

## Reshaping data - long vs wide format

So far, all of the examples we've shown you have been using 'tidy' data. Data is 'tidy' when it follows a couple of rules: *each variable is in its own column*, and *each observation is in its own row*. Making data 'tidy' often means transforming the table from a "wide" format into a "long" format. Long format is efficient to use in data analysis and visualisation and can also be considered "computer readable".

But sometimes when presenting data in tables for humans to read, or when collecting data directly into a spreadsheet, it can be convenient to have data in a wide format. Data is 'wide' when *some or all of the columns are levels of a factor*. An example makes this easier to see.

```{r, message = FALSE}
gbd_wide <- read_csv("global_burden_disease_wide-format.csv")
gbd_long <- read_csv("global_burden_disease_cause-year-sex.csv") 
```

```{r}
head(gbd_wide)

```

```{r}
head(gbd_long)
```

These tables contain the exact same information, but in long (tidy) and wide formats, respectively.

### Pivot values from rows into columns (wider)

If we want to take the long data and put some of the numbers next to each other for easier visualisation, then `pivot_wider()` from the **tidyr** package is the function to do it. It means we want to send a variable into columns, and it needs just two arguments: the variable we want to become the new columns, and the variable where the values currently are.

```{r}
gbd_long %>% 
  pivot_wider(names_from = year, values_from = deaths_millions)
```

This means we can quickly eyeball how the number of deaths has changed from 1990 to 2017 for each cause category and sex. Whereas if we wanted to quickly look at the difference in the number of deaths for females and males, we can change the `names_from =` argument from `= years` to `= sex`. Furthermore, we can also add a `mutate()` to calculate the difference:

```{r}
gbd_long %>% 
  pivot_wider(names_from = sex, values_from = deaths_millions) %>% 
  mutate(Male - Female)
```

All of these differences are positive which means every year, more men die than women. Which make sense, as more boys are born than girls.

And what if we want to look at both `year` and `sex` at the same time? No problem, `pivot_wider()` can deal with multiple variables at the same time, `names_from = c(sex, year)`:

```{r}
gbd_long %>% 
  pivot_wider(names_from = c(sex, year), values_from = deaths_millions)

```

`pivot_wider()` has a few optional arguments that may be useful for you. For example, `pivot_wider(..., values_fill = 0)` can be used to fill empty cases (if you have any) with a value you specified. Or `pivot_wider(..., names_sep = ": ")` can be used to change the separator that gets put between the values (e.g., you may want "Female: 1990" instead of the default "Female_1990"). Remember that pressing F1 when your cursor is on a function opens it up in the Help tab where these extra options are listed.

### Pivot values from columns to rows (longer)

The inverse of `pivot_wider()` is `pivot_longer()`. If you're lucky enough, your data comes from a proper database and is already in the long and tidy format. But if not you'll need to know how to wrangle the variables currently spread across different columns into the tidy format (where each column is a variable, each row is an observation).

`pivot_longer()` can be a little bit more difficult to use as you need to describe all the columns to be collected using a `select_helper`. Run \`?select_helpers and click on the first result in the Help tab for a reminder.

For example, here we want to collect all the columns that include the words Female or Male, the select helper for it is `matches("Female|Male")`:

```{r}
gbd_wide %>% 
  pivot_longer(matches("Female|Male"), 
               names_to = "sex_year", 
               values_to = "deaths_millions") %>% 
  slice(1:6)
```

You're probably looking at the example above and thinking that's all nice and simple on this miniature example dataset, but how on earth will I figure this out on a real-world example. And you're right, we won't deny that `pivot_longer()` is one of the most technically complicated functions in this book, and it can take a lot of trial and error to get it to work. How to get started with your own `pivot_longer()` transformation is to first play with the `select()` function to make sure you are telling R exactly which columns to pivot into the longer format. For example, before working out the `pivot_longer()` code for the above example, we would figure this out first:

```{r}
gbd_wide %>% 
  select(matches("Female|Male"))
```

Then, knowing that `matches("Female|Male")` works as expected inside our little `select()` test, we can copy-paste it into `pivot_longer()` and add the `names_to` and `values_to` arguments. Both of these arguments are new column names that you can make up (in the above example, we are using "sex_year" and "deaths_millions").

### `separate()` a column into multiple columns

While `pivot_longer()` did a great job fetching the different observations that were spread across multiple columns into a single one, it's still a combination of two variables - sex and year. We can use the `separate()` function to deal with that.

```{r}
gbd_wide %>% 
  # same pivot_longer as before
  pivot_longer(matches("Female|Male"), 
               names_to = "sex_year", 
               values_to = "deaths_millions") %>% 
  separate(sex_year, into = c("sex", "year"), sep = "_", convert = TRUE)
```

We've also added `convert = TRUE` to `separate()` so `year` would get converted into a numeric variable. The combination of, e.g., "Female-1990" is a character variable, so after separating them both `sex` and `year` would still be classified as characters. But the `convert = TRUE` recognises that `year` is a number and will appropriately convert it into an integer.

## `arrange()` rows

The `arrange()` function sorts rows based on the column(s) you want. By default, it arranges the tibble in ascending order:

```{r}
gbd_long %>% 
  arrange(deaths_millions) %>% 
  # first 3 rows just for printing:
  slice(1:3)
```

For numeric variables, we can just use a `-` to sort in descending order:

```{r}
gbd_long %>% 
  arrange(-deaths_millions) %>% 
  slice(1:3)
```

The `-` doesn't work for categorical variables; they need to be put in `desc()` for arranging in descending order:

```{r}
gbd_long %>% 
  arrange(desc(sex)) %>% 
  # printing rows 1, 2, 11, and 12
  slice(1,2, 11, 12)
```

### Factor levels

`arrange()` sorts characters alphabetically, whereas factors will be sorted by the order of their levels. Let's make the cause column into a factor:

```{r}
gbd_factored <- gbd_long %>% 
  mutate(cause = factor(cause))
```

When we first create a factor, its levels will be ordered alphabetically:

```{r}
gbd_factored$cause %>% levels()
```

But we can now use `fct_relevel()` inside `mutate()` to change the order of these levels:

```{r}
gbd_factored <- gbd_factored %>% 
  mutate(cause = cause %>% 
           fct_relevel("Injuries"))

gbd_factored$cause %>% levels()
```

`fct_relevel()` brings the level(s) listed in it to the front.

So if we use `arrange()` on `gbd_factored`, the `cause` column will be sorted based on the order of its levels, not alphabetically. This is especially useful in two places:

-   plotting - categorical variables that are characters will be ordered alphabetically (e.g., think barplots), regardless of whether the rows are arranged or not;
-   statistical tests - the reference level of categorical variables that are characters is the alphabetically first (e.g., what the odds ratio is relative to).

However, making a character column into a factor gives us power to give its levels a non-alphabetical order, giving us control over plotting order or defining our reference levels for use in statistical tests.

## Exercises

### Exercise - `pivot_wider()`

Using the GBD dataset with variables `cause`, `year` (1990 and 2017 only), `sex` :

```{r, message = FALSE}
gbd_long <- read_csv("global_burden_disease_cause-year-sex.csv") 
```

Use `pivot_wider()` to put the `cause` variable into columns using the `deaths_millions` as values:

```{r, echo = FALSE}
gbd_long %>% 
  pivot_wider(names_from = cause, values_from = deaths_millions) 


```

**Solution**

```{r, message = FALSE, results = 'hide'}
gbd_long <- read_csv("global_burden_disease_cause-year-sex.csv")
gbd_long %>% 
  pivot_wider(names_from = cause, values_from = deaths_millions)
```

### Exercise - `group_by()`, `summarise()`

Read in the full GBD dataset with variables `cause`, `year`, `sex`, `income`, `deaths_millions`.

```{r, message = FALSE}
gbd_full <- read_csv("global_burden_disease_cause-year-sex-income.csv")

glimpse(gbd_full)
```

Year 2017 of this dataset was shown before but we have multiple years avaiable: `r gbd_full$year %>% unique()`.

Investigate these code examples:

```{r}
summary_data1 <- 
  gbd_full %>% 
  group_by(year) %>% 
  summarise(total_per_year = sum(deaths_millions))

summary_data1

summary_data2 <- 
  gbd_full %>% 
  group_by(year, cause) %>% 
  summarise(total_per_cause = sum(deaths_millions))

summary_data2
```

You should recognise that:

-   `summary_data1` includes the total number of deaths per year.
-   `summary_data2` includes the number of deaths per cause per year.
-   `summary_data1 <-` means we are creating a new dataframe called `summary_data1` and saving (`<-`) results into it. If `summary_data1` was a dataframe that already existed, it would get overwritten.
-   `gbd_full` is the data being sent to the `group_by()` and then `summarise()` functions.
-   `group_by()` tells `summarise()` that we want aggregated results for each year.
-   `summarise()` then creates a new variable called `total_per_year` that sums the deaths from each different observation (subcategory) together.
-   Calling `summary_data1` on a separate line gets it printed.
-   We then do something similar in `summary_data2`.

Compare the number of rows (observations) and number of columns (variables) of `gbd_full`, `summary_data1`, and `summary_data2`.

You should notice that:

-   `summary_data2` has exactly 3 times as many rows (observations) as `summary_data1`. Why?
-   `gbd_full` has 5 variables, whereas the summarised tibbles have 2 and 3. Which variables got dropped? How?

**Answers**

-   `gbd_full` has `r nrow(gbd_full)` observations (rows),
-   `summary_data1` has `r nrow(summary_data1)`,
-   `summary_data2` has `r nrow(summary_data2)`.

`summary_data1` was grouped by year, therefore it includes a (summarised) value for each year in the original dataset. `summary_data2` was grouped by year and cause (`r gbd_full$cause %>% unique()`), so it has 3 values for each year.

The columns a `summarise()` function returns are: variables listed in `group_by()` + variables created inside `summarise()` (e.g., in this case `deaths_peryear`). All others get aggregated.

### Exercise - `full_join()`, `percent()`

For each cause, calculate its percentage to total deaths in each year.

Hint: Use `full_join()` on `summary_data1` and `summary_data2`, and then use `mutate()` to add a new column called `percentage`.

Example result for a single year:

```{r, echo = FALSE}
library(scales)
full_join(summary_data1, summary_data2) %>% 
  mutate(percentage = percent(total_per_cause/total_per_year)) %>% 
  filter(year == 1990)
```

**Solution**

```{r message=FALSE, results='hide'}
library(scales)
full_join(summary_data1, summary_data2) %>% 
  mutate(percentage = percent(total_per_cause/total_per_year)) 
```

### Exercise - `mutate()`, `summarise()`

Instead of creating the two summarised tibbles and using a `full_join()`, achieve the same result as in the previous exercise with a single pipeline using `summarise()` and then `mutate()`.

Hint: you have to do it the other way around, so `group_by(year, cause) %>% summarise(...)` first, then `group_by(year) %>% mutate()`.

Bonus: `select()` columns `year`, `cause`, `percentage`, then `pivot_wider()` the `cause` variable using `percentage` as values.

**Solution**

```{r}
gbd_full %>% 
  # aggregate to deaths per cause per year using summarise()
  group_by(year, cause) %>% 
  summarise(total_per_cause = sum(deaths_millions)) %>% 
  # then add a column of yearly totals using mutate()
  group_by(year) %>% 
  mutate(total_per_year = sum(total_per_cause)) %>% 
  # add the percentage column
  mutate(percentage = percent(total_per_cause/total_per_year)) %>% 
  # select the final variables for better vieweing
  select(year, cause, percentage) %>% 
  pivot_wider(names_from = cause, values_from = percentage)
```

Note that your pipelines shouldn't be much longer than this, and we often save interim results into separate tibbles for checking (like we did with `summary_data1` and `summary_data2`, making sure the number of rows are what we expect and spot checking that the calculation worked as expected).

> R doesn't do what you want it to do, it does what you ask it to do. Testing and spot checking is essential as you will make mistakes. We sure do.

Do not feel like you should be able to just bash out these clever pipelines without a lot of trial and error first.

### Exercise - `filter()`, `summarise()`, `pivot_wider()`

Still working with `gbd_full`:

-   Filter for 1990.

-   Calculate the total number of deaths in the different income groups (`r gbd_full$income %>% unique()`). Hint: use `group_by(income)` and `summarise(new_column_name = sum(variable))`.

-   Calculate the total number of deaths within each income group for males and females. Hint: this is as easy as adding `, sex` to `group_by(income)`.

-   `pivot_wider()` the `income` column.

**Solution**

```{r}
gbd_full %>% 
  filter(year == 1990) %>% 
  group_by(income, sex) %>% 
  summarise(total_deaths = sum(deaths_millions)) %>% 
  pivot_wider(names_from = income, values_from = total_deaths)
```
