---
title: "ML pipeline example R"
execute:
  message: false
format: 
  html: 
    link-external-icon: true
    link-external-newwindow: true
---

1)  Dataset
2)  Splitting into train and test sets
3)  Fitting linear regression algorithm in train
4)  Evaluating algorithm in both train and test sets (compare predctions to true labels through metric RMSE) - <https://yardstick.tidymodels.org/reference/rmse.html>

As an extra:

5)  What happens if now we have a very different dataset? (pandemic has happened and we now have much younger patients coming in). The model no longer holds and we can detect it by comparing the evaluation metrics.

```{r}
#| output: false

# Load necessary libraries
library(tidymodels)
```

```{r}
# Set seed for reproducibility
set.seed(42)
```

```{r}
# Generate initial healthcare data
age <- rnorm(1000, mean = 60, sd = 15)  # Patient age
severity <- runif(1000, 1, 10)          # Illness severity score (1-10)
noise <- rnorm(1000, mean = 0, sd = 2)  # Random noise
LOS <- 1.5 * age + 2 * severity + noise # Linear relationship - normally would not have this! THis is what we are trying to find out!

# Create a data frame
data <- data.frame(Age = age, Severity = severity, LOS = LOS)
```

```{r}
# Split data into training and test sets using tidymodels
data_split <- initial_split(data, prop = 0.8, strata = LOS)
train <- training(data_split)
test <- testing(data_split)

# Define a linear regression model
lm_model <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode('regression')

lm_fit <- lm_model %>% 
          fit(LOS ~ Age + Severity, data = train)

```

```{r}
# Predict on the training and test sets
train_preds <- lm_fit %>%
  predict(new_data = train) %>%
  bind_cols(train)

test_preds <- lm_fit %>%
  predict(new_data = test) %>%
  bind_cols(test)
```

```{r}
# Calculate RMSE for training and test sets
train_rmse <- train_preds %>%
  metrics(truth = LOS, estimate = .pred) %>%
  filter(.metric == "rmse")

test_rmse <- test_preds %>%
  metrics(truth = LOS, estimate = .pred) %>%
  filter(.metric == "rmse")

cat("RMSE on Training Set:", train_rmse$.estimate, "\n")
cat("RMSE on Test Set:", test_rmse$.estimate, "\n")
```

::: important

------------------------------------------------------------------------

EXTRA! We will now create a new dataset, with different underlying model, with data that was not captured in original.
:::

```{r}
# Simulate a more extreme shift in the data distribution (different reality, model building it up)
# Much younger patients with almost negligible severity scores
age_new <- rnorm(200, mean = 20, sd = 5)    # Significantly younger patients
severity_new <- runif(200, 0.1, 1)          # Minimal severity
noise_new <- rnorm(200, mean = 0, sd = 2)
LOS_new <- 0.8 * age_new + 1.3 * severity_new + noise_new #Different underlying model!

# Create a new dataset with highly shifted distribution
new_data <- data.frame(Age = age_new, Severity = severity_new, LOS = LOS_new)
```

```{r}
# Predict on the new data
new_preds <- lm_fit %>%
  predict(new_data = new_data) %>%
  bind_cols(new_data)

# Calculate RMSE for the new, shifted data
new_rmse <- new_preds %>%
  metrics(truth = LOS, estimate = .pred) %>%
  filter(.metric == "rmse")

cat("RMSE on New Data (Highly Shifted Distribution):", new_rmse$.estimate, "\n")
```

```{r}
# Plot: Model Fit on Training Data
ggplot(train_preds, aes(x = LOS, y = .pred)) +
  geom_point(color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(title = "Model Fit on Training Data",
       x = "True LOS (Training)", y = "Predicted LOS") + 
  theme_bw()

# Plot: Model Fit on Test Data
ggplot(test_preds, aes(x = LOS, y = .pred)) +
  geom_point(color = "green") +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(title = "Model Fit on Test Data",
       x = "True LOS (Test)", y = "Predicted LOS") + 
  theme_bw()

# Plot: Model Fit on New Data (Highly Shifted Distribution)
ggplot(new_preds, aes(x = LOS, y = .pred)) +
  geom_point(color = "orange") +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(
       x = "True LOS (New Data)", y = "Predicted LOS") + 
  theme_bw()

```

```{r}
train$Dataset <- "Training"
test$Dataset <- "Testing"
new_data$Dataset <- "Shifted New Data"

# Combine all datasets into one for plotting
combined_data <- rbind(train, test, new_data)

# Plot all datasets
ggplot(combined_data, aes(x = Age, y = Severity, color = LOS)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Dataset) +
  scale_color_viridis_c(option = "plasma") +  # Use a colour scale for LOS
  labs(
    title = "Dataset in Training, Testing, and Validation (new shifted dataset)",
    x = "Age",
    y = "Severity",
    color = "LOS"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

Check out all other models available: <https://www.tidymodels.org/find/parsnip/> Can you try out a different model?
