---
title: 'Python vs R'
format: html
editor: visual
---

Comparison between the main libraries you will be using in R and Python for this module.

------------------------------------------------------------------------

### **`Pandas` Equivalent: `dplyr` from `tidyverse`**

Used for data wrangling. Example: Filtering a dataset.

```{python}
# Python

import pandas as pd
data = pd.DataFrame({'age': [23, 45, 12, 36], 'score': [87, 56, 92, 77]})
filtered_data = data[data['age'] > 20]

filtered_data.head()
```

```{r}
#| output: false
library(dplyr)

data <- data.frame(age = c(23, 45, 12, 36), score = c(87, 56, 92, 77))
filtered_data <- data %>% filter(age > 20)

head(filtered_data)
```

------------------------------------------------------------------------

### **`Scikit-learn` Equivalent: `caret` and `tidymodels`**

Used for machine learning Example: Fitting a linear regression algorithm

```{python}
from sklearn.linear_model import LinearRegression
model = LinearRegression()
X = [[1], [2], [3]]
y = [2, 4, 6]
model.fit(X, y)
print(model.intercept_, model.coef_)
```

```{r}
# R
model <- lm(y ~ x, data = data.frame(x = c(1, 2, 3), y = c(2, 4, 6)))
summary(model)
```

But can also use tidymodels

```{r}
#| output: false
#| 
library(tidymodels)

lm_model <- linear_reg() %>% 
            set_engine('lm') %>% # adds lm implementation of linear regression
            set_mode('regression')

lm_fit <- lm_model %>% 
          fit(y ~ x, data = data.frame(x = c(1, 2, 3), y = c(2, 4, 6)))


lm_fit[["fit"]][["coefficients"]]
```

------------------------------------------------------------------------

### **`Seaborn` Equivalent: `ggplot2`**

Used for plotting Example: Scatter plot

```{python}
# Python
import seaborn as sns
import matplotlib.pyplot as plt
sns.scatterplot(x=[1, 2, 3], y=[2, 4, 6])
plt.show()
```

```{r}
# R
library(ggplot2)

ggplot(data.frame(x = c(1, 2, 3), y = c(2, 4, 6)), aes(x = x, y = y)) +
  geom_point()
```

------------------------------------------------------------------------

### **`Numpy` Equivalent: `base`**

Example: Dot product

```{python}
# Python
import numpy as np
a = np.array([1, 2, 3])
b = np.array([4, 5, 6])
dot_product = np.dot(a, b)
```

```{r}
# R
a <- c(1, 2, 3)
b <- c(4, 5, 6)
dot_product <- sum(a * b)
```

|  |
|:---|
| As an exercise now: |
| 1\) Combine workflow in R - plot and perform linear regression. Can you include the linear regression model in the plot? Check out `geom_smooth` function |
| 2\) Combined workflow in Python - plot and perform linear regression |

Now take your time and explore the machine learning libraries a bit more. If in doubt about anything being learnt in class, these are your best resources:

-   <https://scikit-learn.org/stable/>
-   <https://www.tidymodels.org>
-   <https://topepo.github.io/caret/index.html>
